### 2020-09-26 text/article_vedrovice.Rmd
spatial analysis - combination of vars

output:
    bookdown::pdf_document2:
    # bookdown::word_document2:
        latex: pdflatex
        fig_caption: true
        number_sections: false
        keep_tex: false
bibliography    : "bib_vedrovice.bib"
csl             : "european-journal-of-archaeology.csl"
---

@@

@@

To examine which associations of variables, i.e. artefacts, are infrequent or even not present
and which tend to appear together more often than it would be expected by random chance
the co-occurences of observed and simulated data
are summarized using two statistics suggested by @manly1996.
The *S* statistic is a measure of overall deviation between expected and observed
number of artefact occurences while
the *v* statistic is a measure of difference between co-occurences of
individual artefact with other artifacts
[@manly1996, 475, for details see SI].
To assess relations between groups based on sex, age and origin
statistics introduced by @sosna2008 are used.

@@

Given statistics for independent random process are estimated using Monte Carlo
simulation with 999 iterations.
Although smaller number of permutations should be sufficient to achieve
credible results
[@baddeley2016, 384] we follow the approach taken by @sayer2013 [78].
Ten most extreme realizations of the simulation are discarded as outliers so
the significance level *alpha* is 0.01 [@baddeley2016 410].

@@

## Artifact combinations

By counting co-occurences of analyzed variables on simulated data matrices and comparing
the distributions of the two studied statistics with the values calculated for
the original dataset we can establish whether the overall pattern of artefacts
distribution in the graves is non-random and further identify artefacts that deviate
from the random pattern.
The overall deviation from expected co-occurences, i.e. the *S* statistic
[@manly1996, 475] is significantly large at the 0.05 alpha level.
This suggests that there is strong evidence that the overall distribution of
artefacts in different graves is not random (see Fig. \@ref(fig:sstat)).

Examining distributions of individual artefacts it is found that at the 0.05
significance level only adzes, grinding tools, ceramic bowls and spondylus beads
are distributed randomly.
Close to the 0.05 alpha level are distributions of bone tools and ceramic bottles
(Fig. \@ref(fig:vstat)).
Among the grave goods that are found to be randomly distributed in the graves
across the burial ground are ochre pigment, lithics, pebbles, globular pots and
special types of pottery vessel shapes and spondylus pendants and *buckles*.
The relationships between artefact types frequently occuring and missing together
in graves are visualized in Fig. (\@ref(fig:coocnets)) with
clusters highlighted using fast-greedy algorithm [@newman2004] and strength of the
given association is expressed by the thickness of the edge connecting the nodes.
Artifacts most frequently present with other artifacts are with 5 links grinding tools and
ceramic bowls and with 4 links adzes, non-local lithic tools, ochre pigment and
spondylus beads.
The association is strongest between spondylus beads and (a) ceramic bottles and
(b) pigment; and adzes and (a) non-local lithics and (b) bone tools.
Artifacts that are frequently absent in the graves together are with 6 links
local lithic tools, globular pots and ceramic bowls and spondylus buckels with
5 and 4 links respectively.
The strongest antagonist relationships are between globular pots and (a) bowls and
(b) adzes.

@@

@@

In case of clusters at different percolation distances
there are some deviations from random number of
occurences of given categories at different scales.
Regarding the determination of sex of the bodies, at the threshold of 3.9 meters
we observe that in the largest cluster (nr. 3, Fig. \@ref(fig:perc-plan))
there are more non adults and less male-sexed bodies and in the second largest
cluster (nr. 10) there are slightly more male-sexed bodies than would
be expected under random process.
The overrepresentation of non-adult bodies is repeated in the following
threshold distance of 4.6 meters in the largest cluster (nr. 3).
Last but not least, in the cluster nr. 8 (north of the cemetery) at the distance
of 5.5 meters female-sexed bodies are more frequent than we would expect.
The age categories correspond to the previous obsevations where
most evident is the overrepresentaion of juvenile bodies at the 3.9 meters threshold in
cluster nr. 3.
Exploring the origin of the bodies based on Strontium isotope signature we
find that cluster in the south-western part at threshold distances 3.9 and 5.5 meters
(nr. 10 and 5 respectively) consists of less non-local bodies than would be expected.
Lastly at the distance radius 2.8 meters there are more local bodies in the cenral
cluster of the cemetery (nr. 8).

@@

@@

<!-- randomization -->
<!--
```{r sstat, fig.cap = "Distribution of simulated and observed *S* statistic"}
# knitr::include_graphics(here("plots/s_statistic.pdf"))
```

```{r vstat, fig.cap = "Simulated and observed *V* statistics for different variables"}
# knitr::include_graphics(here("plots", "v_values.pdf"))
```

```{r coocnets, fig.cap = "Networks of variables frequently present together and exluding each other"}
# knitr::include_graphics(here("plots", "cooc_networks.pdf"))
```
-->

<!-- plans -->

@@

```{r label-pp-funs}
label_pp_funs <- paste("Functions *G*, *F*, *K* and *T*.",
"*Dashed line* -- theoretical curve under complete spatial randomness, *solid line* -- empirical curve, *gray area* -- envelope")
```
```{r pp-funs, fig.cap=label_pp_funs}
knitr::include_graphics(here("text/figs/pp_funs.pdf"))
```

@@

<!-- to SI
```{r label-nb-sex}
label_nb_sex <- paste("Mean neighbors in a given sex category.",
"*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
"*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body")
```
```{r nb-sex, fig.cap=label_nb_sex}
knitr::include_graphics(here("text/figs/nb_sex.pdf"))
```

```{r label-nb-age}
label_nb_age <- paste("Mean neighbors in a given age category.",
"*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
"*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body")
```
```{r nb-age, fig.cap=label_nb_age}
knitr::include_graphics(here("text/figs/nb_age.pdf"))
```
-->

@@

```{r perc-distance, fig.cap="Percolation thresholds"}
knitr::include_graphics(here("text/figs/perc_distance.pdf"))
```

@@

>>>>

output:
    bookdown::pdf_document2:
    # bookdown::word_document2:
        latex: pdflatex
        toc: false
        fig_caption: true
        number_sections: false
        keep_tex: false
bibliography    : "bib_vedrovice.bib"
csl             : "european-journal-of-archaeology.csl"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
---

@@

We consider there three variables, sex, age and origin to be indicative of
ones identity in the Neolithic.
To explore the relations between the variables we use them both as individual
categories and combined into following indicator variables -- (1) sex/age,
(2) origin/sex, (3) origin/age and finally (4) origin/sex/age.
Due to frequent missing or indetermined values we decided to limit the studied
sample only to bodies with all of the categories determined and only categories
that occure in more than 5% of burials.

@@

To examine which associations of variables, i.e. artefacts, are infrequent or even not present
and which tend to appear together more often than it would be expected by random chance
the co-occurences of observed and simulated data
are summarized using two statistics suggested by @manly1996.
The *S* statistic is a measure of overall deviation between expected and observed
number of artefact occurences while
the *v* statistic is a measure of difference between co-occurences of
individual artefact with other artefacts
[@manly1996, 475, for details see SI].

To assess relations between groups based on sex, age and origin
statistics introduced by @sosna2008 are used.

@@

Given statistics for independent random process are estimated using Monte Carlo
simulation with 999 iterations.
Although smaller number of permutations should be sufficient to achieve
credible results
[@baddeley2016, 384] we follow the approach taken by @sayer2013 [78].
Ten most extreme realizations of the simulation are discarded as outliers so
the significance level is 0.01 [@baddeley2016 410].

@@

## Artefact combinations

By counting co-occurences of analyzed variables on simulated data matrices and comparing
the distributions of the statistics proposed by @manly1996 with the values calculated for
the original dataset we can establish whether the overall pattern of artefacts
distribution in the graves is non-random and further identify artefacts that deviate
from the random pattern.
The overall deviation from expected co-occurences, i.e. the *S* statistic
[@manly1996, 475] is significantly large at the 0.05 alpha level.
This suggests that there is strong evidence that the overall distribution of
artefacts in different graves is not random. <!--(see Fig. \@ref(fig:sstat))-->

Examining distributions of individual artefacts we find that at the 0.05
significance level only adzes, grinding tools, ceramic bowls and spondylus beads
are distributed randomly.
Close to the 0.05 alpha level are distributions of bone tools and ceramic bottles
(Fig. \@ref(fig:cooc-vstat)).
Among the grave goods that are found to be randomly distributed in the graves
across the burial ground are ochre pigment, lithics, pebbles, globular pots and
special types of pottery vessel shapes and spondylus pendants and buckles.

The relationships between artefact types frequently occuring and missing together
in graves are visualized in Fig. (\@ref(fig:cooc-nets)) with
clusters established using fast-greedy algorithm [@newman2004] and strength of the
given association expressed by the thickness of the edge connecting the nodes.
Artifacts most frequently present with other artifacts are with 5 links grinding tools and
ceramic bowls and with 4 links adzes, non-local lithic tools, ochre pigment and
spondylus beads.
The association is strongest between spondylus beads and (a) ceramic bottles and
(b) pigment; and adzes and (a) non-local lithics and (b) bone tools.
Artifacts that are frequently absent in the graves together are with 6 links
local lithic tools, globular pots and ceramic bowls and spondylus buckels with
5 and 4 links respectively.
The strongest antagonist relationships are between globular pots and (a) bowls and
(b) adzes.

@@

Multitype *K* for combination of age and sex categories suggests clustering process
between bodies determined as adult females and juveniles and vice versa.
The relationship between pairs of non-adults/juveniles is also confirmed.
Several other observed deviations occure at large distances and are same as
in case of multitype *K* for single categories.

@@

The combinations of categories in Delaunay graph show significant relations for
pairs of bodies labeled as (1) females from maturus/senilis category with
juvenile females and (2) unsexed juveniles in general
and as already noted at several places, between (3) pairs of bodies labeled as
juveniles/non-adults.
Gabriel graph in the sex/age category shows relations between (1) adultus female and male bodies,
(2) maturus female and male bodies and (3) maturus female bodies.
The origin/sex combination of categories produces significant deviation from
expected frequencies for (1) non-local female bodies and local non-adults in
Delaunay triangulation and in Gabriel graph for (1) non-local female and local male
bodies and vice versa.
The origin/age combination shows significant departures only in Gabriel graph for
local adultus bodies with (1) non-local maturus bodies and (2) local juvenile bodies.
Last but not least, incorporating all of the categories into the networks we observe
significant results in both of the graphs (1) non-local maturus female bodies and
local juvenile (non-adult) bodies and  in case of Gabrie graph relation
between (2) non-local adultus female and local maturus male bodies.

Clusters formed at different percolation distances
show following deviations from expected numbers of
occurences of given categories at different scales.
Regarding the determination of sex of the bodies, at the threshold of 3.9 meters
we observe that in the largest cluster (nr. 3, Fig. \@ref(fig:perc-plan))
there are more non adults and less male-sexed bodies and in the second largest
cluster (nr. 10) there are slightly more male-sexed bodies than would
be expected under random process.
The overrepresentation of non-adult bodies is repeated in the following
threshold distance of 4.6 meters in the largest cluster (nr. 3).
Last but not least, in the cluster nr. 8 (north of the cemetery) at the distance
of 5.5 meters female-sexed bodies are more frequent than we would expect.
The age categories correspond to the previous obsevations where
most evident is the overrepresentaion of juvenile bodies at the 3.9 meters threshold in
cluster nr. 3.
Exploring the origin of the bodies based on Strontium isotope signature we
find that cluster in the south-western part at threshold distances 3.9 and 5.5 meters
(nr. 10 and 5 respectively) consists of less non-local bodies than would be expected.
Lastly at the distance radius 2.8 meters there are more local bodies in the cenral
cluster of the cemetery (nr. 8).

In the combined categories most of the already observed patterns are repeated and
highlighted.
In the largest cluster across percolation thresholds at 3.9 and 4.6 meters
(nr. 3) there is abundance of (1) local non-adults/juveniles and (2) non-local
female sexed bodies of maturus age category and on the other hand
lesser numbers of (3) juvenile female-sexed bodies.
The central cluster across distances 2.8, 3.9 and 4.6 (nr. 8, 10 and 7 respectively)
shows higher numbers of local male- and female-sexed bodies of adultus category and in case of
female bodies maturus age categories.
The south-western cluster (nr. 10, 7 and 5) at percolation distances 3.9, 4.6 and 5.5
is clearly dominated by local male-sexed bodies of maturus age category.
Last but not least, in case of all the categories combined together a cluster in
the western part of the cemetery at percolation threshold 4.6 and 5.5 (nr. 8 and 6 respectively)
reveals higher counts of non-local female-sexed bodies of adultus age category.

@@

# Conclusions

Our goal was to identify non-random differences between burials both in terms
of the artefact contents of the graves and position within the cemetery.
The burials or more precisely the buried bodies were divided into categories
based on three variables commonly connected to the identity in the Neolithic period.
These categories are biological sex, age and origin.
The exploration was done both for the categories as single variables and for
all of their combinations.

<!-- Artefact combinations -->

<!-- Spatial pattern -->
Spatial organization of the cemetery was explored using point pattern analysis
techniques.
The simpler methods based on cumulative frequency distribution of nearest neighbors
(*G* function) and its generalization for empty space points (*F* function)
did not show deviations from complete spatial random processes.
On the other hand the methods including all the distances between events in the
given area  (*K* and *T* functions) indicated clustering processes are taking place.
*K* function for multivariate point patterns was also used in order to identify
clustering processes between events of different categories.
We consider this not useful in most cases because the distances at which interesting
clustering processes are suggested imply existence of singular cluster across
the cemetery and a finer scale is of much higher interest for our endeavour.

Thus to identify whether close burials are more similar in terms of categories
outlined above two neighborhood graphs, Delaunay triangulation and Gabriel
graph were used.
They are examples of high and low neighbors case scenario respectively.
Most of the observations from the graphs proved not significantly different from
randomization experiment.
There are associations between non-local female-sexed
bodies of maturus age category and (local) juveniles in both of the graphs.
This observation goes in hand with association of female-sexed bodies
and non-adults in Delaunay triangulation.
This is not trivial because in general non-adult bodies are
concentrated in the south-eastern part within the cemetery whereas female-sexed
bodies are scattered all around the area of the cemetery.
Similarly individuals determined as of maturus/senilis age category are in contrast to
juvenile bodies concentrated in the western half of the cemetery.
Furthermore in Gabriel graph there is a clear association between non-local adultus
female-sexed bodies and local maturus male-bodies.
This agrees with general association of female- and male-sexed bodies of
adultus and maturus/senilis age categories in the Gabriel graph.

Most detailed view on the clustering within the cemetery and what kinds of
bodies are prevalent in the given clusters is achieved by the percolation analysis.
The largest cluster across several percolation distances consists of more
non-adult/juvenile bodies and non-local female-sexed bodies of maturus category
on one hand and on the other the male-sexed bodies are underrepresented here.
This is in full accordance with previous observations.
The central cluster visible at small percolation distances showes overrepresentation of
local male- and female-sexed bodies.
Also the south-western cluster has higher counts of local male-sexed bodies of maturus age category and
low amount of non-local bodies.
On the other hand non-local female-sexed bodies of adultus age category are
overrepresented in the western cluster neighboring with the previously mentioned one.

The spatial analysis clearly proved that there is a clustering process taking place
in the cemetery and what more, age, sex and origin of the buried bodies play
structuring roles in the arrangement of the burials.
The patterns are not too strict but some appear to be quite constant across a
range of methods used.

@@

<!-- randomization -->

<!--```{r sstat, fig.cap = "Distribution of simulated and observed *S* statistic"}
# knitr::include_graphics(here("plots/s_statistic.pdf"))
```-->

```{r cooc-vstat, fig.cap = "Simulated and observed *V* statistics for different variables"}
knitr::include_graphics(here("plots", "cooc_vstats.pdf"))
```

```{r label-cooc-nets}
label_cooc_nets <- "Networks of variables frequently present together and exluding each other, non-randomly occuring artefacts in bold"
```
```{r cooc-nets, fig.cap=label_cooc_nets }
knitr::include_graphics(here("text/figs", "cooc_nets.pdf"))
```
<!-- plans -->

```{r plan-sex, fig.cap="Plan of the cemetery, kernel density estimation for differently sexed bodies"}
knitr::include_graphics(here("text/figs/plan_sex.pdf"))
```

```{r plan-age, fig.cap="Plan of the cemetery, kernel density estimation for bodies of different ages"}
knitr::include_graphics(here("text/figs/plan_age.pdf"))
```

```{r plan-orig, fig.cap="Plan of the cemetery, kernel density estimation for bodies of different origins"}
knitr::include_graphics(here("text/figs/plan_origin.pdf"))
```

@@

```{r label-pp-funs}
label_pp_funs <- paste(
  "Functions *G*, *F*, *K* and *T*.",
  "*Dashed line* -- theoretical curve under complete spatial randomness,
  *solid line* -- empirical curve, *gray area* -- envelope")
```
```{r pp-funs, fig.cap=label_pp_funs}
knitr::include_graphics(here("text/figs/pp_funs.pdf"))
```

@@

```{r label-nb-sex}
label_nb_sex <- paste(
  "Mean neighbors in a given category of sex;",
  "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value"
  # "*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body"
)
```
```{r nb-sex, fig.cap=label_nb_sex}
knitr::include_graphics(here("text/figs/nb_sex.pdf"))
```

```{r label-nb-age}
label_nb_age <- paste(
  "Mean neighbors in a given simplified age category"
  # "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
)
```
```{r nb-age, fig.cap=label_nb_age}
knitr::include_graphics(here("text/figs/nb_age.pdf"))
```

```{r label-nb-orig}
label_nb_orig <- paste(
  "Mean neighbors in a given category of origin based on Sr isotopes"
  # "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
  # "*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body"
)
```
```{r nb-orig, fig.cap=label_nb_orig}
knitr::include_graphics(here("text/figs/nb_orig.pdf"))
```

@@

```{r label-perc-dist}
label_perc_dist <- "Normalized sizes of largest clusters at given distances, percolation thresholds labeled"
```
```{r perc-distance, fig.cap=label_perc_dist}
knitr::include_graphics(here("text/figs/perc_distance.pdf"))
```

@@

\pagebreak
